Soundness Proofs for Iterative Deepening

### 2015-11-10 pldi-16/proof/assumptions.tex
start proving saturation for real

>>>>

{\bf Halting.}
From a certain perspective, proving convergence is the same as proving completeness,
which should be impossible for any runtime analysis of a Turing-complete language \cite{entscheidungsproblem}.
However, being already limited by practical real-world CPU budgets,
we are already accepting that many tests will time out.
We are concerned with proving the limited case that when \quicksand~{\em does} terminate with all state spaces completed, the verification is sound.
Inversely, during our proof, when we show that Iterative Deepening will eventually converge to an arbitrary buggy interleaving, we assume that no intermediate state spaces contain nontermination bugs.
If they do, for the sake of testing/verification, we are satisfied with finding that bug instead.
In this proof, we assume that the MC's heuristic infinite loop checker has no false negatives;
i.e., it will never get stuck forever in an infinite loop without identifying a bug, while occasionally producing false positives.
% , such as for a program trying to find a counterexample for Sally's Conjecture.
% TODO: cite some unproved mathematical conjecture by a woman, as an example of an undecidable program?

### 2015-11-10 pldi-16/proof/convergence.tex
start proving saturation for real

%First we will prove a slightly weaker statement, limiting the na\"{i}ve approach's PPs to data races,
%then strengthen it by showing that the all interleavings in the na\"{i}ve state space
%are equivalent to some interleaving made of data-race PPs (and sync API PPs) alone.

%%%%

\newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
\newcommand\ppinstr[1]{\ensuremath{\mathsf{instr}(#1)}}
\newcommand\ppothers[1]{\ensuremath{\mathsf{others}(#1)}}
\newcommand\pppfx[1]{\ensuremath{\mathsf{pfx}(#1)}}
\newcommand\conflicts[2]{\ensuremath{\mathsf{conflicts}(#1,#2)}}
\newcommand\pai{\ensuremath{p_{\alpha{}i}}}
Let
\[
	\mathcal{I} = \{(t_{\alpha{}1}, p_{\alpha{}1}), (t_{\beta{}1}, p_{\beta{}1}), ... (t_{\alpha{}n}, p_{\alpha{}n}), ...\}
\]
be an interleaving trace, where the element $(t_{\alpha{}i}, \pai)$ represents the $i$th transition $t$ of thread $\alpha$, and the associated preemption point $p$.
We will use the following notation:
\begin{itemize}
	\item $\ppnext{p_{\alpha{}i}}$ indicates the next transition by the same thread, $t_{\alpha{}(i+1)}$,
	\item $\ppinstr{p_{\alpha{}i}}$ indicates the first instruction executed during $t_{\alpha{}(i+1)}$. and
	\item $\ppothers{p_{\alpha{}i}}$ indicates the set of all transitions by other threads between $t_{\alpha{}i}$ and $t_{\alpha{}(i+1)}$.
	\item $\conflicts{t_\alpha,t_\beta}$ indicates the set of memory access pairs between $t_\alpha$ and $t_\beta$ such that among each pair, one access belongs to each thread $\alpha$ and $\beta$, the address is the same, and at least one is a write; i.e., precisely the shared memory conflict relation required by DPOR to determine dependent interleavings.
\end{itemize}

@@

% TODO: come up with a better name
\begin{definition}[Relevant preemption point]
	We say a PP $\pai$ is {\em relevant} if either $\ppnext{\pai}$ is a sync API boundary (including a voluntary yield),
	or if
	\[
		\conflicts{\ppothers{\pai}}{\ppinstr{\pai}} \ne \emptyset
	\]
	i.e., the instruction by thread $\alpha$ immediately after $\pai$ has a memory conflict with some other thread interleaved during $\pai$.
\end{definition}
\begin{definition}[Fully-relevant interleaving]
	An interleaving comprised only of relevant PPs.
\end{definition}

@@

\begin{proof}
	Let $\pai$ be the first irrelevant PP of a naive interleaving $\mathcal{I}$.
	We ask, did some thread among $\ppothers{\pai}$ conflict with any instruction from $\ppnext{\pai}$, even though there was no conflict with $\ppinstr{\pai}$?
	\begin{itemize}
			% rho for relevant.
		\item If $\conflicts{\ppothers{\pai}}{\ppnext{\pai}} \ne \emptyset$, or if $t_{\alpha{}i}$ contains a sync API boundary instruction that's not already a PP,
			then let $\rho$ be either the first instruction by $\ppnext{\pai}$ among the conflicts or the first sync API instruction, whichever comes first.
			Let $\pppfx{\rho}$ denote the instruction sequence between $\ppinstr{\pai}$ and $\rho$, including the former but not the latter.
			Now, we output the new interleaving:

			\begin{tabular}{lll}
				\\
				$\mathcal{I}' = \{...,$&$(t_{\alpha{}i} \cup \pppfx{\rho}, \pai'),$& \\
									   &$\ppothers{\pai},$& \\
					       &$(t_{\alpha{}(i+1)}) \setminus \pppfx{\rho}, p_{\alpha{}(i+1)}),$ & $ ...\}$ \\
				\\
			\end{tabular}

			In other words, we have simply reordered $\ppothers{\pai}$ to between $\pppfx{\rho}$ and $\rho$, removing the irrelevant $\pai$ and adding a new PP $\pai'$, which is relevant by the construction of $\rho$.
			We know it must be possible to reorder $\ppothers{\pai}$ to after $\alpha$'s execution because no synchronization is done during $\pppfx{\rho}$, hence the other threads' runnability cannot be affected.
			Likewise $\pppfx{\rho}$ is independent with $\ppothers{\pai}$, so by the soundness of DPOR \cite{dpor}, $\mathcal{I}' \equiv \mathcal{I}$.
		\item Otherwise, $\conflicts{\ppothers{\pai}}{\ppnext{\pai}} = \emptyset$, and no instructions among $\ppnext{\pai}$ are a sync API boundary.
			Then we output the new interleaving:
			\[
				\mathcal{I}' = \{..., (t_{\alpha{}i} \cup t_{\alpha(i+1)}, p_{\alpha(i+1)}), \ppothers{\pai},...\}
			\]
			In other words, we have reordered {\ppothers{\pai}} with the entire next transition by thread $\alpha$.
			We know this must be possible for the same reasons as above, and again, the transitions are independent, so $\mathcal{I}' \equiv \mathcal{I}$.
					       %&$(t_{\alpha{}(i+1)}) \setminus \pppfx{\rho}, p_{\alpha{}(i+1)}),$ & $ ...\}$ \\
	\end{itemize}
	This constitutes an algorithm for inductively converting (or removing) all irrelevant PPs to relevant ones.
\end{proof}

%%%%

@@

\begin{definition}[Reachable data race]
A data race candidate which will be identified by a MC configured to preempt only on locking API boundaries,
or transitively also configured to use data-race PPs of other (already) reachable data races.
\end{definition}

@@

\begin{lemma}
	Given a fully-relevant interleaving
	\[
		\mathcal{I} = \{ ..., (t_{\alpha{}i},\pai), \ppothers{\pai}, (t_{\alpha(i+1)},p_{\alpha(i+1)}), ... \}
	\]
	if $\pai$ is a data-race PP, and the all PPs in the interleaving prefix
	\[
		\mathcal{J} = \{ ..., (t_{\alpha{}i} \cup t_{\alpha(i+1)}',p_{\alpha(i+1)}'), \ppothers(\pai)' \}
	\]
	are reachable, then $\pai$ is reachable.
	\label{lem:coalesce}
\end{lemma}

Note here that in $\mathcal{J}$, some instructions of $t_{\alpha(i+1)}$ may change because they were reordered with $\ppothers(\pai)$;
i.e., $\ppinstr(\pai)$ may not necessarily be the only site of thread communication,
and some other conflicts may cause $t_{\alpha(i+1)}$ to behave differently.
Hence we have labelled it $t_{\alpha(i+1)}'$, and similar for thread $\alpha$'s next PP.
Nevertheless, the first instruction of $t_{\alpha(i+1)}'$ must still be the same as $\ppinstr{\pai}$.

% TODO: What if ppothers changes based on ppinstr(pai)? You say, there must exist some J where ppothers has fewer PPs. You just stop preempting! Those preemptions don't matter. Gotta go fast. Gotta keep the induction metric of getting fewer pps in J than in I.

%%%

\begin{lemma}
	If a bug can be exposed by any fully-relevant interleaving, %some combination of preemptions on any data-race instruction and sync API boundaries,
	Iterative Deepening will eventually test a state space in which all those PPs are enabled\footnote{
		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
	\label{lem:saturation}
\end{lemma}

% TODO: Here's what to do.
% Induct on the PPs by the order of next(my_PP) in the trace.
% Inductive assumption gives that all prior other data race PPs with next(other_PP) before next(my_PP) are reachable,
% Is there a prior data race PP with next(other_PP) *AFTER* next(my_PP)?
% If not, then we are done, and this one is reachable.
% If so, then we ask, which transition among others(other_PP) (this includes us) does instr(other_PP) conflict with?
% <vague handwavy part>
% Then we show that there must be a trace with next(other_PP) interleaved to be before next(my_PP) where, even if the bug doesn't show up, the data race still shows up, which makes other_PP reachable.
% </vague handwavy part>
% Since other_PP is reachable then my_PP is reachable.

\begin{proof}
Let $\mathcal{I} = \{(t_{\alpha{}1}, p_{\alpha{}1}), ...\}$ be the fully-relevant interleaving and PP sequence in the premise of the lemma which exposed a bug.
We proceed by inducting on the PPs according to the order in which for each PP $\pai$, $\ppnext{\pai}$ appeared in the trace.
Note that this is not necessarily the same as the order of the $\pai$s themselves in $\mathcal{I}$;
i.e., it is the order of the racy instructions rather than the order of their immediately preceding preemptions.
%Figure~\ref{fig:why-ppnext} shows why the first PP, ordered by earliest point of preemption, might not necessarily be the first reachable one.
%%% well, the way we do it, might not be the first reachable one either...

\newcommand\pbj{\ensuremath{p_{\beta{}j}}}
For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbj$ with $\ppnext{\pbj}$ occurring before $\ppnext{\pai}$ in $\mathcal{I}$, $\pbj$ is reachable.
% Now to show \pai is reachable.

@@

%%%%

@@

\begin{proof}
By Lemma~\ref{lem:equivalent}, there must be some equivalent fully-relevant interleaving.
%the na\"{i}ve interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption.
Then we simply lean on the soundness of DPOR~\cite{dpor} to guarantee finding the buggy interleaving within this state space.
\end{proof}

>>>>

\newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
\newcommand\ppinstr[1]{\ensuremath{\mathsf{instr}(#1)}}
\newcommand\ppothers[1]{\ensuremath{\mathsf{others}(#1)}}
\newcommand\pppfx[1]{\ensuremath{\mathsf{pfx}(#1)}}
\newcommand\conflicts[2]{\ensuremath{\mathsf{conflicts}(#1,#2)}}
\newcommand\pai{\ensuremath{p_{\alpha{}i}}}
\newcommand\tai{\ensuremath{t_{\alpha{}i}}}
Let
\[
	\mathcal{I} = \{(t_{\alpha{}1}, p_{\alpha{}1}), (t_{\beta{}1}, p_{\beta{}1}), ... (t_{\alpha{}n}, p_{\alpha{}n}), ...\}
\]
be an interleaving trace, where the element $(\tai, \pai)$ represents the $i$th transition $t$ of thread $\alpha$, and the associated preemption point $p$.
We will use the following notation:
\begin{itemize}
	\item $\ppnext{\pai}$ indicates the next transition by the same thread, $t_{\alpha{}(i+1)}$,
	\item $\ppinstr{\pai}$ indicates the first instruction executed during $t_{\alpha{}(i+1)}$. and
	\item $\ppothers{\pai}$ indicates the set of all transitions by other threads between $\tai$ and $t_{\alpha{}(i+1)}$.
	\item $\conflicts{t_\alpha,t_\beta}$ indicates the set of memory access pairs between $t_\alpha$ and $t_\beta$ such that among each pair, one access belongs to each thread $\alpha$ and $\beta$, the address is the same, and at least one is a write; i.e., precisely the shared memory conflict relation required by DPOR to determine dependent interleavings.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{definition}[Relevant preemption point]
	We say a PP $\pai$ is {\em relevant} if either $\ppnext{\pai}$ is a sync API boundary (including a voluntary yield),
	or if
	\[
		\conflicts{\ppothers{\pai}}{\ppinstr{\pai}} \ne \emptyset
	\]
	i.e., the instruction by thread $\alpha$ immediately after $\pai$ has a memory conflict with some other thread interleaved during $\pai$.
\end{definition}
\begin{definition}[Fully-relevant interleaving]
	An interleaving comprised only of relevant PPs.
\end{definition}

@@

\begin{proof}
	Let $\pai$ be the first irrelevant PP of a naive interleaving $\mathcal{I}$.
	We ask, did some thread among $\ppothers{\pai}$ conflict with any instruction from $\ppnext{\pai}$, even though there was no conflict with $\ppinstr{\pai}$?
	\begin{itemize}
			% rho for relevant.
		\item If $\conflicts{\ppothers{\pai}}{\ppnext{\pai}} \ne \emptyset$, or if $\tai$ contains a sync API boundary instruction that's not already a PP,
			then let $\rho$ be either the first instruction by $\ppnext{\pai}$ among the conflicts or the first sync API instruction, whichever comes first.
			Let $\pppfx{\rho}$ denote the instruction sequence between $\ppinstr{\pai}$ and $\rho$, including the former but not the latter.
			Now, we output the new interleaving:

			\begin{tabular}{lll}
				\\
				$\mathcal{I}' = \{...,$&$(\tai \cup \pppfx{\rho}, \pai'),$& \\
									   &$\ppothers{\pai},$& \\
					       &$(t_{\alpha{}(i+1)}) \setminus \pppfx{\rho}, p_{\alpha{}(i+1)}),$ & $ ...\}$ \\
				\\
			\end{tabular}

			In other words, we have simply reordered $\ppothers{\pai}$ to between $\pppfx{\rho}$ and $\rho$, removing the irrelevant $\pai$ and adding a new PP $\pai'$, which is relevant by the construction of $\rho$.
			We know it must be possible to reorder $\ppothers{\pai}$ to after $\alpha$'s execution because no synchronization is done during $\pppfx{\rho}$, hence the other threads' runnability cannot be affected.
			Likewise $\pppfx{\rho}$ is independent with $\ppothers{\pai}$, so by the soundness of DPOR \cite{dpor}, $\mathcal{I}' \equiv \mathcal{I}$.
		\item Otherwise, $\conflicts{\ppothers{\pai}}{\ppnext{\pai}} = \emptyset$, and no instructions among $\ppnext{\pai}$ are a sync API boundary.
			Then we output the new interleaving:
			\[
				\mathcal{I}' = \{..., (\tai \cup t_{\alpha(i+1)}, p_{\alpha(i+1)}), \ppothers{\pai},...\}
			\]
			In other words, we have reordered {\ppothers{\pai}} with the entire next transition by thread $\alpha$.
			We know this must be possible for the same reasons as above, and again, the transitions are independent, so $\mathcal{I}' \equiv \mathcal{I}$.
					       %&$(t_{\alpha{}(i+1)}) \setminus \pppfx{\rho}, p_{\alpha{}(i+1)}),$ & $ ...\}$ \\
	\end{itemize}
	This constitutes an algorithm for inductively converting (or removing) all irrelevant PPs to relevant ones.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{definition}[Reachable data race]
A data race candidate which will be identified by a MC configured to preempt only on locking API boundaries,
or transitively also configured to use data-race PPs of other
%already
reachable data races.
\end{definition}

@@

%%% weird backwards-induction thingie i came up with at the airport that doesn't work
%\begin{lemma}
%	Given a fully-relevant interleaving
%	\[
%		\mathcal{I} = \{ ..., (\tai,\pai), \ppothers{\pai}, (t_{\alpha(i+1)},p_{\alpha(i+1)}), ... \}
%	\]
%	if $\pai$ is a data-race PP, and the all PPs in the interleaving prefix
%	\[
%		\mathcal{J} = \{ ..., (\tai \cup t_{\alpha(i+1)}',p_{\alpha(i+1)}'), \ppothers(\pai)' \}
%	\]
%	are reachable, then $\pai$ is reachable.
%	\label{lem:coalesce}
%\end{lemma}
%
%Note here that in $\mathcal{J}$, some instructions of $t_{\alpha(i+1)}$ may change because they were reordered with $\ppothers(\pai)$;
%i.e., $\ppinstr(\pai)$ may not necessarily be the only site of thread communication,
%and some other conflicts may cause $t_{\alpha(i+1)}$ to behave differently.
%Hence we have labelled it $t_{\alpha(i+1)}'$, and similar for thread $\alpha$'s next PP.
%Nevertheless, the first instruction of $t_{\alpha(i+1)}'$ must still be the same as $\ppinstr{\pai}$.

\begin{lemma}
	All PPs of any fully-relevant interleaving are reachable.
	\label{lem:saturation}
\end{lemma}

% TODO: Here's what to do.
% Induct on the PPs by the order of next(my_PP) in the trace.
% Inductive assumption gives that all prior other data race PPs with next(other_PP) before next(my_PP) are reachable,
% Is there a prior data race PP with next(other_PP) *AFTER* next(my_PP)?
% If not, then we are done, and this one is reachable.
% I

\newcommand\pbh{\ensuremath{p_{\beta{}h}}}
\newcommand\paj{\ensuremath{p_{\alpha{}j}}}
\newcommand\pbk{\ensuremath{p_{\beta{}k}}}
%\newcommand\tak{\ensuremath{t_{\alpha{}k}}}
%\newcommand\pgk{\ensuremath{p_{\gamma{}k}}}
\newcommand\tbh{\ensuremath{t_{\beta{}h}}}
\newcommand\tbk{\ensuremath{t_{\beta{}k}}}
\newcommand\coalesce[1]{\ensuremath{\mathsf{coalesce}(#1)}}

\begin{proof}
Let $\mathcal{I} = \{(t_{\alpha{}1}, p_{\alpha{}1}), ...\}$ be the fully-relevant interleaving and PP sequence in the premise of the lemma which exposed a bug.
We proceed by inducting on the PPs according to the order of their preemptions.
(Note that this is not necessarily the same as the order of the racing instructions $\ppinstr\pai$, which occur in $\ppnext\pai$, not in \tai.)
For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbh$ with $\tbh \prec \tai \in \mathcal{I}$, $\pbh$ is reachable.
We must show that a data race involving $\ppinstr\pai$
is reachable, and we are not allowed to use $\pai$ until we find the data race between $\ppinstr\pai$ and $\ppothers\pai$.

Consider the alternate interleaving prefix:
\[
	\mathcal{I}' = \{..., (\tai \cup \ppnext\pai', \paj), \ppothers\pai' \}
\]
where we have reordered $\ppnext\pai$ to before the execution of $\ppothers\pai$.
Here, $\paj$ is the first sync API PP in $\alpha$ after $\pai$ (as $p_{\alpha(i+1)}$ may be a not-yet-reachable data race PP),
and $\ppnext\pai'$ may include transitions of $\alpha$ beyond $\ppnext\pai$ itself.
%and for brevity, we will abbreviate the union of transitions as $\tak$.
Then, $\ppothers\pai'$ are the other threads' transitions from our target interleaving.
except they may be altered by the presence of some {\em other} data race in $\ppnext\pai'$ occurring after $\ppinstr\pai$.
Likewise, $\ppnext\pai'$ may be altered by some other data race in $\ppothers\pai'$.
The only certainty so far is that $\ppnext\pai'$ begins with $\ppinstr\pai$.

It would be straightforward to show that any other data races in $\ppnext\pai'$, which would be discovered in this interleaving,
could be reordered to after $\ppothers\pai'$,
eventually transforming $\ppnext\pai'$ to contain no conflicting memory accesses but $\ppinstr\pai$ itself.
Unfortunately, $\mathcal{I}'$ is not reachable, as $\ppothers\pai'$ still includes any data-race PPs from $\ppothers\pai$,
which were ordered after $\pai$ in $\mathcal{I}$ and hence not covered by the inductive assumption.
Hence, we must perform the same ``coalescing'' for each thread in $\ppothers\pai'$ that we did for $\alpha$,
which we will abbreviate:
\[
	\mathcal{I}'' = \{..., (\tai \cup \ppnext\pai', \paj), \coalesce{\ppothers\pai'} \}
\]
For each other thread $\beta$, $\coalesce{\ppothers\pai'}$ will contain a single transition,
starting with the first instruction of $\beta$'s corresponding transition in $\ppothers\pai$,
and ending with the next sync API PP, which we'll call $\pbk$.
It follows from sync API PPs being reachable by definition,
and from the inductive hypothesis (which covers the prefix indicated by ``$...$''),
that $\mathcal{I''}$ is reachable.

%%%%%%%%%%%%%%%%% this didn't work out
%We proceed by inducting on the PPs according to the order in which for each PP $\pai$, $\ppnext{\pai}$ appeared in the trace.
%Note that this is not necessarily the same as the order of the $\pai$s themselves in $\mathcal{I}$;
%i.e., it is the order of the racy instructions rather than the order of their immediately preceding preemptions\footnote{
%Why use this strange ordering, rather than simply inducting by the order of preemptions?
%Figure~\ref{fig:why-ppnext} shows why the first PP, ordered by earliest point of preemption, might not necessarily be the first reachable one.
%It is also true that the $\ppnext{\pai}$ order does not necessarily match the PPs' reachability dependence relation, but it is easier to cover the exception cases using this ordering rather than vice versa.}.
%
%\begin{figure}[t]
%	TODO - example of why ppnext
%	\caption{TODO}
%	\label{fig:why-ppnext}
%\end{figure}
%%%% well, the way we do it, might not be the first reachable one either...
%
%\newcommand\pbj{\ensuremath{p_{\beta{}j}}}
%\newcommand\pgk{\ensuremath{p_{\gamma{}k}}}
%For both the base case and inductive step, we know (vacuously, for the former) that for each other PP $\pbj$ with $\ppnext{\pbj} \prec \ppnext{\pai} \in \mathcal{I}$, $\pbj$ is reachable.
%To show $\pai$ is reachable, the proof has two parts.
%First, we must handle any preceding data-race PPs $\pgk$ with $\ppnext{\pai} \prec \ppnext{\pgk}$.
%Second, we must show that when {\pai} is removed, a data-race including $\ppinstr{\pai}$ will be identified.
%%%%%%%%%%%%%%%%%%%

% TODO

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{proof}
By Lemma~\ref{lem:equivalent}, there must be some equivalent fully-relevant interleaving.
%the na\"{i}ve interleaving must be equivalent to some interleaving built of data-race PPs and sync API PPs.
By Lemma~\ref{lem:saturation}, Iterative Deepening will eventually discover and enable all necessary data-race PPs, with sync API PPs being enabled by the maximal state space assumption\footnote{
		In the paper we noted that when a bug is found, \quicksand~cancels all superset state spaces, anticipating that they would find the same bug.
		Because we are concerned with verifying the absence of bugs, we say this also suffices for convergence, but for brevity, we will leave this condition implicit.}.
Then we simply lean on the soundness of DPOR~\cite{dpor} to guarantee finding the buggy interleaving within this state space.
\end{proof}
